Constructing a universe for the setoid model

### 2020-07-08 paper.tex
minor

Induction-recursion is arguably a nicer and more natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$) \cite{malatasta13smallir}.

@@

$\U$ and $\El$ can be given by $\U :\equiv \Sigma (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.

@@

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \Set_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

What is left now is to define the setoid equality relation on the universe, as
well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
%
Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
equivalence relation on their respective sets $\El\ A$ and $\El\ B$. Intuitively,
since elements of a setoid are only ever compared to elements of the same
setoid, this should only be possible if $A$ and $B$ are codes for the same setoid,
that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
relations is expressed via the type $\inUU$ just defined, hence we would
expect $A \sim_{\setoidU} B$ to be defined as follows:
\[
(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma (R: A \to B \to \mProp)(\inUU\ p_A\ p_B\ R)
\]

@@

The full definition of the universe is concluded with the remaining definitions,
like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
more or less straighforwardly. The final result does not use
induction-recursion, but it's nevertheless an instance of infinitary
induction-induction. The ability to define arbitrary, infinitary
inductive-inductive types clashes, again, with our objective of keeping the
metatheory as small and simple as possible. The next step is therefore to reduce
this inductive-inductive universe to one that does not require (infinitary)
induction-induction.

@@

We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
\Sigma (\GG_0 : \textsf{Con}_0)(\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
:\equiv \Sigma(A_0 :
\textsf{Ty}_0)(\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
is straighforward:
\begin{alignat*}{10}
  & \bullet && :\equiv (\bullet_0 , \bullet_1) \\
  & (\GG_0 , \GG_1) , (A_0 , A_1) && :\equiv ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
  & \iota\ (\GG_0 , \GG_1) && :\equiv (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && :\equiv
    (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
\end{alignat*}

>>>>

Induction-recursion is arguably a nice and natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$) \cite{malatasta13smallir}.

@@

$\U$ and $\El$ can be given by $\U :\equiv \Sigma\ (A : \mType)\ (\textsf{in-U}\ A)$ and $\El :\equiv \pi_1$.

@@

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) :\equiv \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}
\\

@@

What is left now is to define the setoid equality relation on the universe, as
well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
%
Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
equivalence relation on their respective sets $\El\ A$ and $\El\ B$. Intuitively,
since elements of a setoid are only ever compared to elements of the same
setoid, this should only be possible if $A$ and $B$ are codes for the same setoid,
that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
relations is expressed via the type $\inUU$ just defined, hence we would
expect $A \sim_{\setoidU} B$ to be defined as follows:
\[
(A, p_A) \sim_{\setoidU} (B, p_B) :\equiv \Sigma\ (R: A \to B \to \mProp)\ (\inUU\ p_A\ p_B\ R)
\]

@@

The full definition of the universe is concluded with the remaining definitions,
like $\reflu, \reflel$, etc., which can be adapted from their IR counterparts
more or less straightforwardly. The final result does not use
induction-recursion, but it's nevertheless an instance of infinitary
induction-induction. The ability to define arbitrary, infinitary
inductive-inductive types clashes, again, with our objective of keeping the
metatheory as small and simple as possible. The next step is therefore to reduce
this inductive-inductive universe to one that does not require (infinitary)
induction-induction.

@@

We can recover the original inductive-inductive type as $\textsf{Con} :\equiv
\Sigma\ (\GG_0 : \textsf{Con}_0)\ (\textsf{Con}_1\ \GG_0)$ and $\textsf{Ty}\ \GG
:\equiv \Sigma\ (A_0 :
\textsf{Ty}_0)\ (\textsf{Ty}_1\ (\pi_1\ \GG)\ A_0)$. Recovering the constructors
is straightforward:
\begin{alignat*}{10}
  & \bullet && :\equiv (\bullet_0 , \bullet_1) \\
  & (\GG_0 , \GG_1) , (A_0 , A_1) && :\equiv ((\GG_0 \mathbin{,_0} A_0) , (\GG_1 ,_1 A_1)) \\
  & \iota\ (\GG_0 , \GG_1) && :\equiv (\iota_0\ \GG_0 , \iota_1\ \GG_1) \\
  & \Pi\ \{\GG_0 , \GG_1\}(A_0 , A_1) (B_0 , B_1) && :\equiv
    (\Pi_0\ \GG_0\ A_0\ B_0 , \Pi_1\ \GG_1\ A_1\ B_1)
\end{alignat*}
